\Sof{TBA - New Titles in the comments}

### 2020-05-25 paper/icsme20.tex
intro/abstract

The main contributions of the present work are:
%
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.
	\item A replication package with all scripts and data created to perform the
	empirical evaluation, for reproducibility. Available online:
  \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%-
This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in the
protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and discusses their
implications; section~\ref{sec:threats} enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%

\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software code base. However, these patches may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security patches impact
the maintainability of software.

@@

%%%%%%% 
Software maintainability is designed as the degree to which an application is understood, 
repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is 
approximately $75\%$ of the cost related to a project. To answer the following three 
research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
to measure the impact of security patches on the maintainability of open-source software. 
%

@@

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security refactorings that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
% \end{itemize}
%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
previous work focused on comparing the impact of security refactorings with
regular refactorings on maintainability.
We analyze the maintainability of regular commits (i.e., commits not related
to security fixes) and use them as a baseline.
The baseline dataset uses the security commits dataset as input.
Both datasets have the same size. For each
security commit, a commit with an approximate size is searched. Due to the complexity 
of some patches, it was not possible to find patches with the exact number of added and 
deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. 
The algorithm integrates the following steps:

@@

\textbf{Guidelines/Metrics:} Each patch performs a set of changes
on the source code of the software. These changes may have a different
impact on the guidelines/metrics used to measure software maintainability. 
Figure~\ref{fig:guidelines} shows the impact of security patches 
on each guideline individually and the average impact on all guidelines 
together - $M(v)$. Under each guideline is stated the 
metric used to the calculations. Table~\ref{tab:guidelines} describes 
in more detail the metrics behind the guidelines presented. 
For each type of guideline, three types of impact of patches in 
the software maintainability are presented: 
\emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). 
For each type of guideline, a boxplot 
and a swarmplot are presented to show the variability and dipersion of
the results alongside the number of absolute
and relative cases of each impact. Next to each
type of guideline, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
average impact of all guidelines.

Regarding the impact of security patches for each guideline, we observe
that overall patches have a more positive impact than negative in
all guidelines/metrics. However, patching vulnerabilities also has a very
significant negative impact on all guidelines - between $10\%$ and $40\%$. 
\emph{Write Short Units of 
Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
most affected guidelines. 
This may imply that developers when patching 
vulnerabilities fail to design/implement solutions that continue to respect 
the limit bounds of branch points (/programming conditions) and function/module 
sizes (e.g., functions should not have more than $15$ LOCs) that are recommended
for coding best practices. Still, on 
respecting bound limits, developers also seem to not consider the limit 
of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
violated when the patch requires to input new information to a function/class 
and developers fail to use the \emph{Introduce Parameter Object} refactoring 
pattern. Results do not provide statistical significance to the \emph{Separate 
Concerns in Modules} guideline, i.e., results should be read carefully.

@@

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) of the cases have negative impact
on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7$)
have a negative impact. The larger number of negative cases may be 
explained by guidelines with higher concentration of negative cases with higher amplitudes
such as WShortUC, WSimpleUC and SCM. In other words, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
ESLint, Infer, and more) should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While static code analysis
may be daunting due to the number of rules and different effects
on maintainability, BCH only uses the top-$10$ of the guidelines
with the highest effect on maintainability. Static analysis 
can be used to complement the source code analysis, introduce
the capability of vulnerability detection, and, consequently,
reduce the risk of hindering maintainability.

@@

\emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
is a type of mistake that, in proper conditions, could contribute to the 
introduction of vulnerabilities within that product. The CWE enumeration
system was used to classify the vulnerabilities included in our study.
Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
the impact of all patches grouped using the first-level weaknesses from
the \emph{Research Concepts} list on CWE; and, the remaining charts
present the impact on maintainability of the sub-weaknesses from two of
the weaknesses presented in \emph{(a)}.

>>>>

The main contributions of the present work are:
%
\begin{itemize}
	\item An empirical study on the impact of security changes on software
  maintainability and what patterns need more attention.
  \item \Rui{results, conclusions}
	\item A replication package with all scripts and data created to perform the
	empirical evaluation, for reproducibility. Available online:
  \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%-
This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in the
protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and discusses their
implications; section~\ref{sec:threats} enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%

\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. \Rui{strange sentence -->} It turns out that fixing these issues is as simple as
modifying the software code base. However, these patches may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security patches impact
the maintainability of software.

@@

%%%%%%% 
Software maintainability is \Rui{designed??} designed as the degree to which an application is understood, 
repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is 
approximately $75\%$ of the cost related to a project. To answer the following three 
research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
to measure the impact of security patches on the maintainability of open-source software. 
%

@@

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security refactorings that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
% \end{itemize}
%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
previous work focused on comparing the impact of security refactorings with
regular refactorings on maintainability.
We analyze the maintainability of regular commits (i.e., commits not related
to security fixes) and use them as a baseline.
\Rui{strange sentence:} The baseline dataset uses the security commits dataset as input.
Both datasets have the same size. For each
security commit, a commit with an approximate size is fetched. Due to the complexity 
of some patches, it was not possible to find patches with the exact number of added and 
deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. 
The algorithm integrates the following steps:

@@

\textbf{Guidelines/Metrics:} Each patch performs a set of changes
on the source code of the software. These changes may have a different
impact on the guidelines/metrics used to measure software maintainability. 
Figure~\ref{fig:guidelines} shows the impact of security patches 
on each guideline individually and the average impact on all guidelines 
together - $M(v)$. \Rui{doesn't parse -->} Under each guideline is stated the 
metric used to the calculations. Table~\ref{tab:guidelines} describes 
in more detail the metrics behind the guidelines. 
For each type of guideline, three types of impact of patches in 
the software maintainability are presented: 
\emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). 
For each type of guideline, a boxplot 
and a swarmplot are presented to show the variability and dipersion of
the results alongside the number of absolute
and relative cases of each impact. Next to each
type of guideline, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
average impact of all guidelines.

\Rui{the next couple of sentences read contradictory...}
Regarding the impact of security patches for each guideline, we observe
that overall patches have a more positive impact than negative in
all guidelines/metrics. However, patching vulnerabilities also has a very
significant negative impact on all guidelines - between $10\%$ and $40\%$. 
\emph{Write Short Units of 
Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and 
\emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
most affected guidelines. 
\Rui{The next sentence should be in a morer  abstract way.. too much detail, no?}
This may imply that developers when patching 
vulnerabilities fail to design/implement solutions that continue to respect 
the limit bounds of branch points (/programming conditions) and function/module 
sizes (e.g., functions should not have more than $15$ LOCs) that are recommended
for coding best practices. Still, on 
respecting bound limits, developers also seem to not consider the limit 
of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
violated when the patch requires to input new information to a function/class 
and developers fail to use the \emph{Introduce Parameter Object} refactoring 
pattern. Results do not provide statistical significance to the \emph{Separate 
Concerns in Modules} guideline, i.e., results should be read carefully.

@@

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) of the cases have negative impact
on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7$)
have a negative impact. The larger number of negative cases may be 
explained by guidelines with higher concentration of negative cases with higher amplitudes
such as \Rui{these acronyms came out of the blue.} WShortUC, WSimpleUC and SCM. In other words, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
ESLint, Infer, and more) should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While static code analysis
may be daunting due to the number of rules and different effects
on maintainability, BCH only uses the top-$10$ of the guidelines
with the highest effect on maintainability. Static analysis 
can be used to complement the source code analysis, introduce
the capability of vulnerability detection, and, consequently,
reduce the risk of hindering maintainability.

@@

\emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
is a type of mistake that, \Rui{meaning?} in proper conditions, could contribute to the 
introduction of vulnerabilities within that product. The CWE enumeration
system was used to classify the vulnerabilities included in our study.
Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
the impact of all patches grouped using the first-level weaknesses from
the \emph{Research Concepts} list on CWE; and, the remaining charts
present the impact on maintainability of the sub-weaknesses from two of
the weaknesses presented in \emph{(a)}.
